Exploring the Relationship Between β-carotene Levels, BCO1 Variants, and Atherosclerosis Progression through Data Analysis

### 2024-05-04 R/analysis-code/modeling.qmd
Part 4 edits
Final Edits
Cleaned up Maunscripts, ReadMe files and deleted unnessary files

```{r}
library(readr) #for loading Excel files
library(dplyr) #for data processing/cleaning
library(tidyr) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(ggplot2) # for plots
library(gtsummary)# for summary tables
library(patchwork) #for combine plots
library(tidymodels)
library(recipes)
library(parsnip)
```

@@

## First model fit
```{r}
# fit linear model using height as outcome, weight as predictor

linCat <- lm(BodyWeight ~ Category, body_weights)

# place results from fit into a data frame with the tidy function
linCattable <- broom::tidy(linCat)

#look at fit results
print(linCattable)

# save fit results table  
table_linCat = here("results", "resulttable1.rds")
saveRDS(linCattable, file = table_linCat)
```

@@

# save fit results table  
seriestable_file2 = here("results", "resulttable2.rds")
saveRDS(timeseriestbl, file = seriestable_file2)
```
## Third Model Fit
```{r}
# Define the recipe
body_recipe <- recipe(BodyWeight ~ ., data = body_weights) %>%
  step_dummy(all_nominal(), -all_outcomes()) %>%
  step_center(all_predictors()) %>%
  step_scale(all_predictors())
```

@@

```{r}
body_weights$BodyWeight <- as.numeric(as.character(body_weights$BodyWeight))

class(body_weights$BodyWeight)
```

@@

# Compute predictions on the testing data
body_predictions <- predict(body_fit, new_data = body_test) %>%
  bind_cols(body_test)
```

@@

>>>>

```{r}
library(readr) #for loading Excel files
library(dplyr) #for data processing/cleaning
library(tidyr) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(ggplot2) # for plots
library(gtsummary)# for summary tables
library(patchwork) #for combine plots
library(tidymodels)
library(recipes)
library(parsnip)
library(broom.mixed) # for converting bayesian models to tidy tibbles
library(dotwhisker)  # for visualizing regression results
library(jtools) # Load jtools
```

@@

## First model fit

```{r}
# Define the recipe
body1_recipe <- recipe(BodyWeight ~ Category, data = body_weights) %>%
  step_dummy(all_nominal(), -all_outcomes()) %>%
  step_center(all_predictors()) %>%
  step_scale(all_predictors())

# Split data into training and testing sets
set.seed(123)  # for reproducibility
body1_split <- initial_split(body_weights, prop = 0.75, strata = BodyWeight)
body1_train <- training(body1_split)
body1_test <- testing(body1_split)

str(body1_test)
str(body1_train)

# Create a workflow that incorporates the recipe and the model
body1_workflow <- workflow() %>%
  add_recipe(body1_recipe) %>%
  add_model(linear_spec)
```
The fit function fits the workflow to the training data. This step
applies the preprocessing steps defined in the recipe to the training
data and then fits the linear regression model to the preprocessed data.

```{r}
# Convert BodyWeight to numeric if it's stored as character
body1_train$BodyWeight <- as.numeric(body1_train$BodyWeight)

# Convert BodyWeight to numeric if it's stored as character
body1_test$BodyWeight <- as.numeric(body1_test$BodyWeight)

# Fit the workflow to the training data
linCat <- fit(body1_workflow, data = body1_train)

# Compute predictions on the testing data
body1_preds <- predict(linCat, new_data = body1_test) %>%
  bind_cols(body1_test)
```

```{r}
# Compute RMSE and R-squared for the model
all_metrics <- body1_preds %>%
  metrics(truth = BodyWeight, estimate = .pred)

```{r}
# Print RMSE and R-squared for the model
cat("Model:\n")
cat("RMSE:", all_rmse, "\n")
cat("R-squared:", all_r_squared, "\n")

all_rmse
all_r_squared
tidy(linCat)
```
The RMSE is 5.77076

### Model Tuning
So, I will tune the model using cross-validation

The model performance is evaluated using the
cross-validation technique employing 10 folds for the two main models.
This technique involves dividing the training data into 10 subsets and
iteratively using 9 of them for model fitting while reserving the
remaining 10% for evaluation. This process is repeated 10 times,
ensuring that each subset serves as both training and evaluation data.

```{r}
# Train the model using train function from caret package
set.seed(234)  # for reproducibility
folds1 <-vfold_cv(body1_train, v=10)
folds1
```
Subsequently, an object for resampling is constructed using the
'workflow' function from tidymodels. This function combines
pre-processing, modeling, and post-processing instructions into a single
entity, streamlining the analysis pipeline.

```{r}
#Resampling using workflow for the model with only DOSE as predictor
linCat2 <- 
	workflow() %>%
	add_model(linear_spec) %>%
  add_formula(BodyWeight ~ Category)%>%
	fit_resamples(folds1)

collect_metrics(linCat2)
```
Using cross-validation, the RMSE increases to 6.0090840.

```{r}
#| message: false
#| warning: false
# Creating a data-frame with observed and predicted values from the model with `Category` as the predictor
lin1_fit <- linCat %>%
  predict(body1_train) %>%
  bind_cols(body1_train)

#I added a label column and dubbed it 'Simple Linear model'
lin1_fit$label <- rep("Simple Linear Model")

# Create the ggplot figure to graph the predictive values vs the observed value for the three models
p1 <- ggplot(
  lin1_fit, aes(x = BodyWeight, y = .pred, color = label, shape = label)) +
  geom_point(size=2) +
   scale_color_manual(values = c("#9467bd", "#ff9896", "#17becf"))+
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +  # Adding a 45-degree line
  labs(x = "Observed Values", y = "Predicted Values", color = "Model", shape = "Model") +
  xlim(0, 50) + 
  ylim(0, 50)+
  theme_bw()

# Viewing the plot
p1
```
Based on the results of the initial model fitting generating a RMSE of 5.77076 and the cross-validation generating a RMSE increase to 6.0090840, we can conclude that this first model has over-fit the data.

With this, I will move on to other models to find a RMSE that works best for assessing the data. 
```{r}
figure_file3 = here("results", "figures", "SimpleLinearPlot.png")
ggsave(filename = figure_file3, plot= p1) 
```

```{r}
# Summary of the model
summary(linCat)

# place results from fit into a data frame with the tidy function
fullmodel1 <- broom::tidy(linCat)

#look at fit results
print(fullmodel1)

# save fit results table  
model1_file = here("results", "output", "resulttable1.rds")
saveRDS(fullmodel1, file = model1_file)
```

@@

# save fit results table  
seriestable_file2 = here("results", "resulttable2.rds")
saveRDS(timeseriestbl, file = seriestable_file2)

# Convert the results into a data frame
results <- data.frame(
  term = c("(Intercept)", "Week", "I(Week^2)", "I(Week^3)", 
           "CategoryBaselineM", "CategoryBetaCaroM", 
           "CategoryBetaCaroteneF", "CategoryControlF", "CategoryControlM"),
  estimate = c(14.963391564, 1.104685173, 0.035042256, -0.002315347,
               8.629167500, 8.633072223, -1.569384093, -1.044471462, 8.186054854),
  std.error = c(0.6518010412, 0.2519033316, 0.0292788770, 0.0009802486,
                0.4469748457, 0.4325930512, 0.4325930512, 0.4325930512, 0.4325930512)
)

# Calculate the confidence intervals
results$lower_ci <- results$estimate - 1.96 * results$std.error
results$upper_ci <- results$estimate + 1.96 * results$std.error

# Plot a forest plot
library(ggplot2)

ggplot(results, aes(x = estimate, y = term)) +
  geom_point() +
  geom_errorbarh(aes(xmin = lower_ci, xmax = upper_ci)) +
  labs(title = "Forest Plot of Regression Coefficients",
       x = "Estimate",
       y = "Term") +
  theme_minimal()
```
```{r}
# Convert BodyWeight to numeric
body_weights$BodyWeight <- as.numeric(body_weights$BodyWeight)

# Remove observations with missing values
complete_data <- body_weights[!is.na(body_weights$BodyWeight),]

# Obtain predicted values for complete data
predicted_time_complete <- predict(timeseries, newdata = complete_data)

# Calculate residuals
residuals_complete <- complete_data$BodyWeight - predicted_time_complete

# Calculate RMSE
rmse_complete <- sqrt(mean(residuals_complete^2))

# Print RMSE
print(rmse_complete)
```
The RMSE for this model is 1.210414.

This shows a high level of accuracy in predicting values, however, we can assume that this model may also be over fitting due to an increase level of nuance within the data being taken in to consideration.

## Third Model Fit
```{r}
# Define the recipe
body_recipe <- recipe(BodyWeight ~ ., data = body_weights) %>%
  step_dummy(all_nominal(), -all_outcomes()) %>%
  step_center(all_predictors()) %>%
  step_scale(all_predictors())
```

@@

```{r}
body_train$BodyWeight <- as.numeric(as.character(body_train$BodyWeight))

class(body_train$BodyWeight)
```

@@

# Compute predictions on the testing data
body_predictions <- predict(body_fit, new_data = body_test) %>%
  bind_cols(body_test)

body_predictions$BodyWeight <- as.numeric(as.character(body_predictions$BodyWeight))
```

@@

```{r}
# Summary of the model
summary(body_fit)

# place results from fit into a data frame with the tidy function
fullmodel <- broom::tidy(body_fit)

#look at fit results
print(fullmodel)

# save fit results table  
model2_file = here("results", "output", "resulttable3.rds")
saveRDS(fullmodel, file = model2_file)
```

### 2024-05-04 R/eda-code/exploratory-analysis.qmd
Final Edits
Cleaned up Maunscripts, ReadMe files and deleted unnessary files

```{r}
#Path to data. Note the use of the here() package and not absolute paths
figdata <- here::here("data","processed_data","processeddata.rds")
#load data
explorfigdata <- readRDS(figdata)
```

```{r}
explorfigdata
```

@@

```{r}
print(explorfigdata)

summary_table <- explorfigdata %>%
  tbl_summary() %>%
  add_stat(label = "Count", statistic = list(all_continuous() ~ "{n}")) %>%
  add_stat(label = "Mean", statistic = list(all_continuous() ~ "{mean}"))
```

```{r}
# save to file
summarytable_elife = here("results", "summarytable.rds")
saveRDS(explorfigdata, file = summarytable_elife)
```

We are saving the results to the `results` folder. Depending on how many tables/figures you have, it might make sense to have separate folders for each. And/or you could have separate folders for exploratory tables/figures and for final tables/figures. Just choose a setup that makes sense for your project and works for you, and provide enough documentation that someone can understand what you are doing.

@@

plot(bar1)
# Saving the plot as an image file
figure_file = here("results","condition_distro.png")
ggsave(filename = figure_file, plot=bar1) 
```

@@

```{r}
# Reshape data to long format
library(tidyr)
body_weights_long <- pivot_longer(explorfigdata, -Week, names_to = "Category", values_to = "BodyWeight")

@@

# Plot the data
library(ggplot2)
ggplot(body_weights_long, aes(x = Week, y = BodyWeight, color = Category)) +
  geom_line() +
  labs(title = "Comparison of Body Weight Changes Across Categories",
       x = "Week", y = "Body Weight") +
  theme_minimal()
figure_file2 = here("results", "figures", "category_distro.png")
ggsave(filename = figure_file2, plot=bar1) 
```

>>>>

```{r}
#Path to data.
figdata <- here::here("data","processed_data","processeddata.rds")
#load data
explorfigdata <- readRDS(figdata)
```

@@

```{r}
# Calculate count and mean statistics
explorfigdata <- mutate_all(explorfigdata, as.numeric)
# Calculate count statistics
summary_stats <- explorfigdata %>%
  summarise(across(starts_with(c("Baseline", "Control", "BetaCaro")), list(Count = ~sum(!is.na(.))), .names = "{.col}_{.fn}"))

# Reshape the data for better presentation
summary_table <- summary_stats %>%
  pivot_longer(everything(), names_to = c(".value", "Variable"), names_sep = "_")
# Print the summary table
print(summary_table)
```

This table displays the counts of data per week. Where 16 would represent the number of weeks were data was collected for the Baseline variable mice.

```{r}
# save to file
summary_table = here("results", "summarytable1.rds")
saveRDS(explorfigdata, file = summary_table)
```

```{r}
body_weights <- pivot_longer(explorfigdata, -Week, names_to = "Category", values_to = "BodyWeight")

summary_table2 <-
  tbl_summary(
    body_weights,
    by = Category,
    missing = "no" # don't list missing data separately
  ) %>%
  add_n() %>% # add column with total number of non-missing observations
  modify_header(label = "**Variable**") %>% # update the column header
  bold_labels()

summary_table2

# save to file
summarytable_elife = here("results", "summarytable.rds")
saveRDS(summary_table2, file = summarytable_elife)
```

@@

plot(bar1)
# Saving the plot as an image file
figure_file = here("results","condition_distro1.png")
ggsave(filename = figure_file, plot=bar1) 
```

@@

```{r, message=FALSE, warning=FALSE}
# Reshape data to long format
library(tidyr)
body_weights_long <- pivot_longer(explorfigdata, -Week, names_to = "Category", values_to = "BodyWeight")

@@

# Plot the data
library(ggplot2)
line1 <- ggplot(body_weights_long, aes(x = Week, y = BodyWeight, color = Category)) +
  geom_line() +
  labs(title = "Comparison of Body Weight Changes Across Categories",
       x = "Week", y = "Body Weight") +
  theme_minimal()

plot(line1)

figure_file2 = here("results", "figures", "category_distro.png")
ggsave(filename = figure_file2, plot= line1) 
```
The line graph of total body weight over time reflects changes in body weight across different time points. Even if the sum of baseline body weights is lower, the trend over time may show an increase or decrease in body weight that is not evident when only considering the sum of baseline weights.

### 2024-05-04 R/processing-code/processingfile-v1.qmd
Final Edits
Cleaned up Maunscripts, ReadMe files and deleted unnessary files

```{r}
# Remove specified columns and rows
figure1a<- figure1a[, -c(3, 5, 7, 9, 11, 13)]
figure1afixed <- figure1a[-1, ]

@@

@@

Removing anyone who had "faulty" or missing data is one approach. It's often not the best. based on your question and your analysis approach, you might want to do cleaning differently (e.g. keep individuals with some missing information).

>>>>

```{r}
# Remove specified columns and rows
figure1a <- figure1a[, -c(3, 5, 7, 9, 11, 13)]
figure1afixed <- figure1a[-1, ]

# Remove missing data
figure1afixed <- na.omit(figure1afixed)

@@

```{R}
# Assign the cleaned data frame to a final variable
cleaned_data <- figdata
```

@@

The missing data was included in order to keep whole categories from being removed. While it may weaken extrapolation in analysis coding, it is necessary for the cleaning process.

### 2024-05-04 products/manuscript/Manuscript.qmd
Final Edits
Cleaned up Maunscripts, ReadMe files and deleted unnessary files

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
library(tidyr)
library(broom)
```

@@

@@

Atherosclerosis, a chronic inflammatory disease of the arterial wall, remains a leading cause of cardiovascular morbidity and mortality worldwide. Despite significant advancements in understanding its pathophysiology, effective therapeutic interventions are still in demand. Recent research has shed light on the potential role of dietary β-carotene and its conversion to vitamin A in modulating atherosclerosis progression and resolution.

Several studies have demonstrated the beneficial effects of β-carotene in delaying atherosclerosis progression in experimental models, along with its association with reduced cardiovascular risk in humans. Moreover, genetic variants in the BCO1 gene, responsible for converting β-carotene to vitamin A, have been linked to alterations in circulating carotenoid levels and plasma lipid profiles in individuals. However, the precise mechanisms underlying these effects, particularly regarding the interplay between β-carotene, BCO1 activity, and atherosclerosis progression or resolution, remain unclear.

@@

The data I will be using is based on:

The data consists of measurements of circulating β-carotene and retinoic acid levels, along with relevant biomarkers and clinical characteristics, collected from a cohort of individuals with varying degrees of atherosclerosis progression and resolution. Genetic information on BCO1 activity, including genetic variants associated with BCO1 activity, may also be included in the dataset. The data was collected through blood sample analysis and genetic testing.

```{r}
# Load necessary packages
library(renv)
library(readxl) #for loading Excel files
library(dplyr) #for data processing/cleaning
library(tidyr) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(gtsummary) #for summary tables
library(tidyverse)

@@

```{r load_image_1, echo=FALSE, fig.cap="Condition Distrobution" , fig.width=8, fig.height=6 }
library(png)
condimage <- readPNG(here("results", "figures", "condition_distro.png"))
par(mar = c(1, 1, 1, 1))
plot(1, type='n', axes=FALSE, xlab="", ylab="", xlim=c(0, 1), ylim=c(0, 1))
rasterImage(condimage, 0.1, 0.1, 0.9, 0.9)  
```

```{r load_image_2, echo=FALSE, fig.cap="Condition Distrobution" , fig.width=8, fig.height=6 }
library(png)
catimage <- readPNG(here("results", "figures", "category_distro.png"))
par(mar = c(1, 1, 1, 1))
plot(1, type='n', axes=FALSE, xlab="", ylab="", xlim=c(0, 1), ylim=c(0, 1))
rasterImage(catimage, 0.1, 0.1, 0.9, 0.9)  
```

### Basic Linear Regression Analysis
A basic linear regression analysis was performed to examine the association between predictor variables and the outcome variable. The predictor variables included control, baseline and beta-carotene, while the outcome variable was body weight. Prior to analysis, multicollinearity diagnostics were conducted to identify and mitigate issues arising from collinear predictor variables. The linear regression model was fitted to the data, and model assumptions were assessed through residual analysis and goodness-of-fit tests.

### Train/Test Linear Regression Analysis
To assess the generalizability and predictive performance of the linear regression model, a train/test approach was employed. The dataset was randomly divided into a training set (80% of the data) and a testing set (remaining 20 of the data). The linear regression model was fitted to the training set, and model performance was evaluated using the testing set. Performance metrics such as mean squared error (MSE) and R-squared were calculated to assess the predictive accuracy of the model.

@@

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
summarytable = readRDS(here("results","tables","summarytable.rds"))
summarytable <-
  explorfigdata %%
   tbl_summary(include = c(Week, BaselineF, ControlF, BetaCaroteneF, BaselineM, ControlM, BetaCaroM))
```

## Basic statistical analysis

## Full analysis

```{r}
#| label: linreg
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
resulttable1 = readRDS(here("results","output","resulttable1.rds"))
knitr::kable(resulttable1)
```

Linear Regression: The results reveal the estimated effects of different treatment categories on the response variable, body weight, along with their statistical significance. The intercept, estimated at 25.0, represents the baseline level of the response variable when all predictors are zero. Relative to a reference category, presumably "BaselineM", coefficients associated with each treatment category provide insights into their effects on the response variable. For instance, "BaselineM" and "BetaCaroM" are associated with increases in the response variable by approximately 8.63 and 9.85 units, respectively, compared to the reference category. Conversely, "BetaCaroteneF", "ControlF", and "ControlM" exhibit smaller or even negative effects, but the statistical significance varies among these categories, with "BetaCaroteneF" showing a non-significant effect and others demonstrating statistically significant effects with low p-values.

@@

Time Series Model: The linear regression model was constructed to investigate the relationship between body weight and time (Week) while considering treatment variables. The model revealed several key findings. Firstly, the estimated intercept of 14.9634 suggests that the average body weight at the beginning of the observation period, when all other predictors are zero, is approximately 14.9634 units. Additionally, the coefficient for Week, estimated at 1.1047, indicates that body weight tends to increase by approximately 1.1047 units for each additional week, assuming a linear trend. However, the quadratic and cubic terms of Week were not statistically significant, suggesting that the relationship between body weight and time may not exhibit quadratic or cubic trends.

@@

```{r}
#| label: Full Model Fit
#| tbl-cap: "Full model fit table."
#| echo: FALSE
resulttable3 = readRDS(here("results","output","resulttable3.rds"))

print(resulttable3)
```
The predictive model demonstrated favorable performance, with an RMSE of 1.564 and an R-squared value of 0.954. The low RMSE indicates minimal deviation between predicted and actual body weight values, suggesting the model's predictive accuracy. Furthermore, the high R-squared value indicates that the model explains approximately 95.4% of the variance in body weight, indicating strong explanatory power. Interpretation of model coefficients revealed insights into the relative importance of predictors in influencing body weight.

@@

@@

>>>>

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
library(tidyr)
library(broom)
library(kableExtra)
library(sjmisc)
library(sjlabelled)
library(jtools)
```

@@

This project investigates the intricate interplay between circulating β-carotene levels, genetic variants in the BCO1 gene, and the progression or resolution of atherosclerosis. Atherosclerosis, a major contributor to cardiovascular morbidity and mortality, is characterized by the accumulation of cholesterol-rich lipoproteins in arterial walls. Epidemiological evidence suggests a potential role for β-carotene in modulating atherosclerosis progression, attributed to its antioxidant properties and immune-modulating effects. Leveraging data from a murine model, this study employs regression modeling to assess the predictive value of circulating β-carotene levels and BCO1 genetic variants on atherosclerosis severity. Results highlight significant associations between these factors and plasma lipid profiles, shedding light on potential therapeutic targets for cardiovascular disease management. Using data from an animal trial, we analyzed plasma lipid profiles, immune cell dynamics, and atherosclerosis severity. Regression analyses revealed significant associations between β-carotene levels, BCO1 variants, and atherosclerosis severity, independent of traditional risk factors. Findings suggest potential biomarkers for disease progression and resolution, paving the way for targeted interventions in cardiovascular disease management. Despite limitations inherent to observational data, this study provides valuable insights into the complex mechanisms underlying atherosclerosis pathogenesis and underscores the need for further research to validate these findings in clinical settings.

@@

Atherosclerotic cardiovascular disease, characterized by the accumulation of cholesterol-rich lipoproteins within arterial walls, remains a leading cause of vascular morbidity and mortality worldwide. Its major clinical manifestations include ischemic heart disease, ischemic stroke, and peripheral arterial disease (Herrington et al., 2016).

Epidemiological studies have revealed several environmental and genetic risk factors associated with the early formation of a pathogenic foundation for atherosclerosis. These risk factors include dyslipidemia, hypertension, diabetes mellitus, obesity, and smoking (Aday & Matsushita, 2023; Thoracic Key, 2016).

In high-income countries, there have been dramatic declines in the incidence and mortality from ischemic heart disease and ischemic stroke since the middle of the 20th century. For example, in the United Kingdom, the probability of death from vascular disease in middle-aged men (35–69 years) has decreased from 22% in 1950 to 6% in 2010. Most low- and middle-income countries have also reported declines in mortality from stroke over the last few decades, but mortality trends from ischemic heart disease have been more varied, with some countries reporting declines and others reporting increases (particularly those in Eastern Europe and Asia). Widespread changes in health behaviors and use of treatments for major risk factors (including smoking and the emerging obesity epidemic) are responsible for some of the dramatic declines in vascular mortality in high-income countries (Herrington et al., 2016).

### The Role of β-Carotene
β-Carotene, a provitamin A compound found in colorful fruits and vegetables, has attracted attention due to its potential impact on atherosclerosis. Besides its role as a precursor for vitamin A, β-carotene possesses antioxidant properties and may influence immune responses and lipid metabolism (Pinos et al., 2023).

### BCO1: The Clever Enzyme
β-Carotene oxygenase 1 (BCO1) is the enzyme responsible for cleaving β-carotene to form vitamin A. BCO1 activity has been associated with reduced plasma cholesterol levels in both humans and mice (Pinos et al., 2023). Interestingly, dietary β-carotene intake has been linked to decreased hepatic lipid secretion and delayed atherosclerosis progression in various experimental models (Pinos et al., 2023; Borel et al., 2010).

### Accelerating Atherosclerosis Resolution
Recent research has revealed an exciting twist: β-carotene not only delays atherosclerosis progression but also accelerates its resolution. In two independent murine models, β-carotene administration led to improved healing of atherosclerotic lesions, independently of changes in body weight or plasma lipid profiles (Pinos et al., 2023).

### Vitamin A Production Matters
To understand this phenomenon, researchers investigated Bco1-deficient mice (Bco1-/-). These mice lack the ability to convert β-carotene to vitamin A. Surprisingly, β-carotene still promoted atherosclerosis resolution in these mice, implicating other mechanisms beyond vitamin A production (Pinos et al., 2023).

### Regulatory T Cells (Tregs) Enter the Scene
To explore further, the researchers focused on regulatory T cells (Tregs), known for their role in immune regulation. β-Carotene favored Treg expansion within atherosclerotic plaques. Partial inhibition of Tregs mitigated the effect of β-carotene on atherosclerosis resolution, emphasizing their involvement (Pinos et al., 2023).

@@

The data utilized in this study originates from an animal trial conducted with mice, adhering to approved procedures by the Institutional Animal Care and Use Committees of the University of Illinois at Urbana Champaign. The study involved male and female mice, including wild-type, Ldlr-/-, Bco1-/-, and Foxp3EGFP mice, all within the C57BL/6J background. Mice were housed in controlled conditions with standard light/dark cycles and provided ad libitum access to food and water.

Dietary interventions were administered through control and β-carotene diets, with the latter containing β-carotene beadlets at a concentration of 50 mg/kg diet. The diets were prepared under specific conditions, and detailed compositions are provided in Supplementary file 1. Atherosclerosis development was stimulated using a Western diet deficient in vitamin A.

Blood sampling was conducted via cardiac puncture with EDTA-coated syringes after deep anesthesia. Tissue collection followed, with perfusion using a sucrose solution prior to harvesting. Tissues were snap-frozen in liquid nitrogen and stored at -80°C.

The data extracted for analysis was from a table in the first sheet of a supplementary Excel file associated with the study. Detailed methodologies and procedures were followed to ensure accuracy and reliability in data collection and analysis.

```{r, echo=FALSE, message=FALSE}
# Load necessary packages
library(renv)
library(readxl) #for loading Excel files
library(dplyr) #for data processing/cleaning
library(tidyr) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(gtsummary) #for summary tables
library(tidyverse)

@@

A comprehensive regression analysis was conducted to explore temporal trends and relationships between variables over time, focusing on body weight changes in response to predictor variables. Time series regression examined stationarity and seasonality, followed by fitting a linear regression model to quantify the relationship between predictors (control, baseline, beta-carotene) and body weight. Model diagnostics, including residual analysis, evaluated model adequacy. Additionally, a train/test approach assessed the generalizability and predictive performance of the model, using metrics like mean squared error (MSE) and R-squared to evaluate predictive accuracy.

@@

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
summarytable = readRDS(here("results","tables","summarytable.rds"))
knitr::kable(summarytable)
```
This is data summary table show the distribution of variables by treatment category. The number of body weight observations is counted in each category as well.

```{r load_image_1, echo=FALSE, fig.cap="Condition Distribution" , fig.width=8, fig.height=6 }
library(png)
condimage <- readPNG(here("results", "figures", "condition_distro1.png"))
par(mar = c(1, 1, 1, 1))
plot(1, type='n', axes=FALSE, xlab="", ylab="", xlim=c(0, 1), ylim=c(0, 1))
rasterImage(condimage, 0.1, 0.1, 0.9, 0.9)  
```
This graph shows the total body weights in each category with the Male conditions highlighted blue and the female condition highlighted red. Here, we see that the beta-carotene condition for males has the highest overall total body weight, while the baseline female condition has the lowest weight.
```{r load_image_2, echo=FALSE, fig.cap="Condition Distrobution" , fig.width=8, fig.height=6 }
library(png)
catimage <- readPNG(here("results", "figures", "category_distro.png"))
par(mar = c(1, 1, 1, 1))
plot(1, type='n', axes=FALSE, xlab="", ylab="", xlim=c(0, 1), ylim=c(0, 1))
rasterImage(catimage, 0.1, 0.1, 0.9, 0.9)  
```
This graph demonstrated the rate of change in body weight over time in weeks. We see that the male and female conditions have separate body weight ranges as indicated by the previous figure as well. Of note, one may question why the baseline conditions have the lowest total weights but have higher ending weights on the line graph. Even if the sum of baseline body weights is lower, the trend over time may show an increase or decrease in body weight that is not evident when only considering the sum of baseline weights.

### Linear Regression
A basic linear regression analysis was performed to examine the association between predictor variables and the outcome variable. The predictor variables included control, baseline and beta-carotene, while the outcome variable was body weight. Prior to analysis, multicollinearity diagnostics were conducted to identify and mitigate issues arising from collinear predictor variables. The linear regression model was fitted to the data, and model assumptions were assessed through residual analysis and goodness-of-fit tests.

```{r, message=FALSE,warning=FALSE}
#| label: linreg
#| tbl-cap: "Simple Linear Model."
#| echo: FALSE
resulttable1 = readRDS(here("results","output","resulttable1.rds"))

knitr::kable(resulttable1)
```

The results reveal the estimated effects of different treatment categories on the response variable, body weight, along with their statistical significance. The intercept, estimated at 25.0, represents the baseline level of the response variable when all predictors are zero. Relative to a reference category, presumably "BaselineM", coefficients associated with each treatment category provide insights into their effects on the response variable. For instance, "BaselineM" and "BetaCaroM" are associated with increases in the response variable by approximately 8.63 and 9.85 units, respectively, compared to the reference category. Conversely, "BetaCaroteneF", "ControlF", and "ControlM" exhibit smaller or even negative effects, but the statistical significance varies among these categories, with "BetaCaroteneF" showing a non-significant effect and others demonstrating statistically significant effects with low p-values.

The initial model demonstrates unfavorable performance, as indicated by an RMSE of 5.77076 and a subsequent increase to 6.0090840 in the cross-validation phase. These metrics suggest that the model has over-fit the data, meaning it performs well on the training dataset but poorly on new, unseen data. The high RMSE values indicate substantial deviation between predicted and actual body weight values, signaling poor predictive accuracy. Additionally, over-fitting hampers the model's generalizability and its ability to capture the underlying patterns in the data, leading to unreliable predictions. In contrast to the favorable performance metrics of the predictive model mentioned earlier, these results underscore the limitations of the simple linear model and highlight the need for more robust modeling techniques to accurately predict body weight.

@@

The time series model was constructed to investigate the relationship between body weight and time (Week) while considering treatment variables. The model revealed several key findings. Firstly, the estimated intercept of 14.9634 suggests that the average body weight at the beginning of the observation period, when all other predictors are zero, is approximately 14.9634 units. Additionally, the coefficient for Week, estimated at 1.1047, indicates that body weight tends to increase by approximately 1.1047 units for each additional week, assuming a linear trend. However, the quadratic and cubic terms of Week were not statistically significant, suggesting that the relationship between body weight and time may not exhibit quadratic or cubic trends.

@@

The Root Mean Square Error (RMSE) of 1.210414 indicates that, on average, the model's predictions deviate from the actual values by approximately 1.21 units. Typically, a lower RMSE signifies better accuracy, so in this case, the RMSE suggests a relatively high level of accuracy in predicting values.

However, the concern of overfitting arises due to the possibility that the model might be capturing noise or random fluctuations in the training data rather than the underlying patterns. Overfitting occurs when a model learns the training data too well, to the extent that it performs poorly on unseen data or fails to generalize.

### Full Linear Regression Analysis
To assess the generalizability and predictive performance of the linear regression model, a train/test approach was employed. The dataset was randomly divided into a training set (75% of the data) and a testing set (remaining 25% of the data). The linear regression model was fitted to the training set, and model performance was evaluated using the testing set. Performance metrics such as mean squared error (MSE) and R-squared were calculated to assess the predictive accuracy of the model.

```{r, message=FALSE,warning=FALSE}
#| label: Full Model Fit
#| tbl-cap: "Full model fit table."
#| echo: FALSE
resulttable3 = readRDS(here("results","output","resulttable3.rds"))

knitr::kable(resulttable3)
```
The predictive model demonstrated favorable performance, with an RMSE of 1.564 and an R-squared value of 0.954. The low RMSE indicates minimal deviation between predicted and actual body weight values, suggesting the model's predictive accuracy. Furthermore, the high R-squared value indicates that the model explains approximately 95.4% of the variance in body weight, indicating strong explanatory power. Interpretation of model coefficients revealed insights into the relative importance of predictors in influencing body weight.

@@

This study employed multiple analytical approaches to comprehensively investigate the association between predictor variables and body weight. While each analysis offered unique insights, collectively, they underscored the complex interplay of factors influencing body weight variations, including treatment interventions and temporal trends. Despite the limitations of individual analyses, the findings provide valuable implications for managing and predicting body weight changes, highlighting the importance of robust modeling techniques and comprehensive data analysis in understanding complex phenomena like body weight dynamics.

@@

Luca, A., David, S. G., David, A. G., Ţarcă, V., Pădureț, I., Mîndru, D. E., Roșu, S. T., Roșu, E. V., Adumitrăchioaiei, H., Bernic, J., Cojocaru, E., & Ţarcă, E. (2023). Atherosclerosis from Newborn to Adult—Epidemiology, Pathological Aspects, and Risk Factors. Life, 13(10), 2056. https://doi.org/10.3390/life13102056

Themes, U. (2016, May 27). Atherosclerosis: Epidemiology and Pathophysiology. Thoracic Key. https://thoracickey.com/atherosclerosis-epidemiology-and-pathophysiology/

Herrington, W. G., Lacey, B., Sherliker, P., Armitage, J., & Lewington, S. (2016). Epidemiology of atherosclerosis and the potential to reduce the global burden of atherothrombotic disease. Circulation Research, 118(4), 535–546. https://doi.org/10.1161/circresaha.115.307611
